---
title: "ObjectiveRisksPUMR"
author: "Calvin Isch"
date: "09/05/2022"
output: html_document
---

# Load data
user_responses.csv contains relevant variables from longitudinal US cohort. 
GBD_risks.csv contains GBD data relevant for each participant

# Updated variables:
Income, savings, assets -- from 1-9 value to middle of bracket
Wealth -- sum of assets and savings
PUMR -- 100 - likelihood of living to 79 with max effort 3 times (March, May, July)
PCMR -- likelihood of living to 79 with max effort - likelihood with min effort, same times
PUMRCause -- likelihood of dying before 79 with max effort to avoid specific cause
HealthBehavior -- sum of 6 health risk behaviors

``` {r , echo=FALSE}
# Load data
df <- read.csv("user_responses.csv")
df_sev <- read.csv("GBD_risks.csv")

library(moments)
library(lme4)
library(bestNormalize)



# Change to middle of brackets for income
wealth_vals = c(7500,20000,30000,42500,62500,87500,125000,175000,250000)
wealth_levels <- function(x){
  if (is.na(x)){
    result <- NA
  }
  else{
    result <- wealth_vals[x]
  }
  return(result)
}

wealth_vals = c(7500,20000,30000,42500,62500,87500,125000,175000,250000)
df$Income2 <- as.numeric(lapply(df$Income, wealth_levels))

# Change to middle of brackets for savings and assets
wealth_vals = c(250,2750,12500,32500,57500,85000,125000,175000,250000)
df$Savings2 <- as.numeric(lapply(df$Savings, wealth_levels))
df$Assets2 <- as.numeric(lapply(df$Assets, wealth_levels))
df$Wealth <- df$Savings2 + df$Assets2


pumr <- function(x){
  if (is.na(x)){
    result <- NA
  }
  else {
    result <- 100 - x
  }
  return(result)
}


df$PUMRMarch <-  as.numeric(lapply(df$EffortLive.7, pumr))
df$PUMRMay <-  as.numeric(lapply(df$EffortLive.9, pumr)) 
df$PUMRJuly <-  as.numeric(lapply(df$EffortLive.11, pumr))

df$PCMRMarch <- df$EffortLive.7 - df$NoEffortLive.7
df$PCMRMay <- df$EffortLive.9 - df$NoEffortLive.9
df$PCMRJuly <- df$EffortLive.11 - df$NoEffortLive.11

# Do not subtract these from 100 because they are perceived likelihood of dying rather than living
df$PUMRCovid <-  as.numeric(df$Mortality1)
df$PUMRAir <-  as.numeric(df$Mortality2)
df$PUMRWater <-  as.numeric(df$Mortality3)
df$PUMRViolence <-  as.numeric(df$Mortality4)
df$PUMRTravel <-  as.numeric(df$Mortality5)
df$PUMROccupation <-  as.numeric(df$Mortality6)
df$PUMRDisaster <-  as.numeric(df$Mortality7)
df$PUMRDisease <-  as.numeric(df$Mortality8)

# Sum health risk behavior DOSPERT scale questions
df$HealthBehavior <- as.numeric(df$RiskBehavior1 + df$RiskBehavior2 + df$RiskBehavior3 + df$RiskBehavior4 + df$RiskBehavior5 + df$RiskBehavior6)

# Two rows were duplicated in the original dataset. Remove these.
df <- df[-c(271,439),] 

# Count of people who answer each question
colSums(!is.na(df))
# Remove other null values from state
colSums(!is.na(df[which(df$State_Name != ""),]))[8]

```



# Study 1: Environmental risk SEV and PUMR
Mixed effects model exploring if air pollution SEV is predictive of PUMR for air pollution when controlling for education, wealth, income, age, sex, and State
``` {r Study 1}
good_summary <- function(lov){
  s = round(sd(lov,na.rm=TRUE),2)
  m = round(mean(lov,na.rm=TRUE),2)
  med = round(median(lov,na.rm=TRUE),2)
  sk = round(skewness(lov,na.rm=TRUE),2)
  k = round(kurtosis(lov,na.rm=TRUE),2)
  return(paste('Mean:',m,'SD:',s,'Median:',med,'Skew:',sk,'Kurtosis:',k,sep=" "))
} 

# Model 4 from the pre-reg = Study 1 in the paper.
df_model <- merge(df,df_sev[df_sev[,'Risk'] == 'Environmental/occupational risks',],by=c('User'))
print(paste('SEV Vals -- ',good_summary(df_model$X5_year_range)))
print(paste('PUMR -- ',good_summary(df_model$PUMRMarch)))

# Test which of the control variables are significantly related to the outcome variable
df_model <- df_model [df_model$Sex %in% c("Male", "Female"), ]
wilcox.test(df_model$PUMRMarch~df_model$Sex)
cor.test(df_model$Wealth,df_model$PUMRMarch,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Education,df_model$PUMRMarch,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Income2,df_model$PUMRMarch,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Age,df_model$PUMRMarch,use="pairwise.complete.obs",method='spearman')

# Normalize wealth, education, income, age (the significant control), PUMR, and SEV
df_model <- merge(df,df_sev[df_sev[,'Risk'] == 'Environmental/occupational risks',],by=c('User'))
df_model$PUMRMarchNormal <- bestNormalize(df_model$PUMRMarch)$x.t
df_model$X5_year_rangeNormal <- bestNormalize(df_model$X5_year_range)$x.t
df_model$AgeNormal <- bestNormalize(df_model$Age)$x.t
df_model$WealthNormal <- bestNormalize(df_model$Wealth)$x.t
df_model$IncomeNormal <- bestNormalize(df_model$Income2)$x.t

risk_mixed_normal = lmer(PUMRMarchNormal ~ X5_year_rangeNormal + WealthNormal + Education + IncomeNormal + AgeNormal + (1 | State_Name), data = df_model)

summary(risk_mixed_normal)

# This produces an intercept for state of 0, so the random variable isn't helping any. We can remove and use a more basic model

risk_normal = lm(PUMRMarchNormal ~ X5_year_rangeNormal + WealthNormal + Education + IncomeNormal + AgeNormal, data = df_model) 

summary(risk_normal,diagnostics=TRUE)

# Creates Supplemental Fig. S1
png("diagnostic_model1.png", width=800, height=600)
par(mfrow = c(2, 2))
plot(risk_normal)
dev.off()


df_model2 <- df_model[complete.cases(df_model[,c("X5_year_range","Wealth","PUMRMarch","Education","Income2","Age","Sex") ]),]

table(df_model2$Sex)
good_summary(df_model2$Age)
good_summary(df_model2$Income2)
good_summary(df_model2$Wealth)
good_summary(df_model2$Education)


```


``` {r Study 2}
library(ggcorrplot)

df_model <- df[c("PUMRMarch","PUMRCovid","PUMRAir","PUMRWater","PUMRViolence","PUMRTravel","PUMROccupation","PUMRDisaster","PUMRDisease")]

# Spearman correlation because so many non-normal
corr <- cor(df_model, method = "spearman", use = "complete.obs")

ggcorrplot(corr, hc.order = TRUE, type = "lower",
   lab = TRUE)

```

# Regression between specific risks and PUMR
``` {r Study 2}
df_model <- df

# Test which of the control variables are significantly related to the outcome variable
df_model <- df_model [df_model$Sex %in% c("Male", "Female"), ]
wilcox.test(df_model$PUMRMarch~df_model$Sex)
cor.test(df_model$Wealth,df_model$PUMRMarch,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Education,df_model$PUMRMarch,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Income,df_model$PUMRMarch,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Age,df_model$PUMRMarch,use="pairwise.complete.obs",method='spearman')

df_model <- df

df_model$PUMRMarchNormal <- bestNormalize(df_model$PUMRMarch)$x.t
df_model$AgeNormal <- bestNormalize(df_model$Age)$x.t
df_model$WealthNormal <- bestNormalize(df_model$Wealth)$x.t
df_model$IncomeNormal <- bestNormalize(df_model$Income)$x.t
df_model$PUMRCovidNormal <- bestNormalize(df_model$PUMRCovid)$x.t
df_model$PUMRAirNormal <- bestNormalize(df_model$PUMRAir)$x.t
df_model$PUMRWaterNormal <- bestNormalize(df_model$PUMRWater)$x.t
df_model$PUMRViolenceNormal <- bestNormalize(df_model$PUMRViolence)$x.t
df_model$PUMRTravelNormal <- bestNormalize(df_model$PUMRTravel)$x.t
df_model$PUMROccupationNormal <- bestNormalize(df_model$PUMROccupation)$x.t
df_model$PUMRDisasterNormal <- bestNormalize(df_model$PUMRDisaster)$x.t
df_model$PUMRDiseaseNormal <- bestNormalize(df_model$PUMRDisease)$x.t

risk_mixed_normal = lmer(PUMRMarchNormal ~  PUMRCovidNormal+ PUMRAirNormal + PUMRWaterNormal + PUMRViolenceNormal + PUMRTravelNormal + PUMROccupationNormal + PUMRDisasterNormal + PUMRDiseaseNormal +WealthNormal + Education + IncomeNormal + AgeNormal + (1 | State_Name), data = df_model)

summary(risk_mixed_normal)

risk_normal <- lm(PUMRMarchNormal ~  PUMRCovidNormal+ PUMRAirNormal + PUMRWaterNormal + PUMRViolenceNormal + PUMRTravelNormal + PUMROccupationNormal + PUMRDisasterNormal + PUMRDiseaseNormal +WealthNormal + Education + IncomeNormal + AgeNormal, data = df_model)

summary(risk_normal)


# check the residuals
png("my_plot2.png", width=6, height=6)
par(mfrow = c(2, 2))
plot(risk_normal)
dev.off()

# PUMR for disease generally and education are only significant predictors of PUMR
good_summary(df_model$PUMRCovid)
good_summary(df_model$PUMRAir)
good_summary(df_model$PUMRWater)
good_summary(df_model$PUMRViolence)
good_summary(df_model$PUMRTravel)
good_summary(df_model$PUMROccupation)
good_summary(df_model$PUMRDisaster)
good_summary(df_model$PUMRDisease)
good_summary(df_model$Wealth)
good_summary(df_model$Education)
good_summary(df_model$Income2)
good_summary(df_model$Age)
good_summary(df_model$PUMRMarch)

```


# Stability of PUMR over time
``` {r Study 3}
library(ggplot2)

df_model <- df[c("PUMRMarch","PUMRMay","PUMRJuly")]


# None of these appear normal
good_summary(df_model$PUMRMarch)
good_summary(df_model$PUMRMay)
good_summary(df_model$PUMRJuly)

cor.test(df_model$PUMRMarch,df_model$PUMRMay,use="pairwise.complete.obs",method='pearson')
cor.test(df_model$PUMRMarch,df_model$PUMRJuly,use="pairwise.complete.obs",method='pearson')
cor.test(df_model$PUMRMay,df_model$PUMRJuly,use="pairwise.complete.obs",method='pearson')

# Might be interesting to control for education, SES, and age.
ggplot(df_model, 
       aes(x=PUMRMarch, y=PUMRMay,col='May')) +
  geom_point(color='#D55E00') + 
  geom_smooth(method=lm,color='#D55E00') + 
  ylab('PUMR Later Test') + 
  geom_point(aes(x=PUMRMarch, y=PUMRJuly),color='#009e73') + 
  geom_smooth(aes(x=PUMRMarch, y=PUMRJuly,col='July'),method=lm,color='#009e73') + 
  theme_classic() + scale_color_brewer(palette="Dark2")

wilcox.test(df$PUMRMarch, df$PUMRMay, paired = TRUE, alternative = "two.sided",na.rm =TRUE)
wilcox.test(df$PUMRMarch, df$PUMRJuly, paired = TRUE, alternative = "two.sided",na.rm =TRUE)
```



``` {r Study 4 - Health behavior}

# DOSPERT =================================================
print(paste('Health Behavior -- ',good_summary(df$HealthBehavior)))
print(paste('PUMR -- ',good_summary(df$PUMRMarch)))

df_model <- df
# Test which of the control variables are significantly related to the outcome variable
cor.test(df_model$Wealth,df_model$HealthBehavior,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Education,df_model$HealthBehavior,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Income,df_model$HealthBehavior,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Age,df_model$HealthBehavior,use="pairwise.complete.obs",method='spearman')
df_model <- df_model [df_model$Sex %in% c("Male", "Female"), ]
wilcox.test(df_model$HealthBehavior~df_model$Sex)

# Wealth, age, and sex are significant
df_model$HealthBehaviorNormal <- bestNormalize(df_model$HealthBehavior)$x.t
df_model$PUMRMarchNormal <- bestNormalize(df_model$PUMRMarch)$x.t
df_model$AgeNormal <- bestNormalize(df_model$Age)$x.t
df_model$WealthNormal <- bestNormalize(df_model$Wealth)$x.t

risk_mixed_normal = lm(HealthBehaviorNormal~ PUMRMarchNormal + WealthNormal +  AgeNormal + Sex, data = df_model)
summary(risk_mixed_normal)


# OTHER HEALTH ===================================
df_health <- read.csv('health_behave.csv')
df_model <- merge(df,df_health,by=c('User'))
good_summary(df_model$EffortHealth.13)

# Test which of the control variables are significantly related to the outcome variable
cor.test(df_model$Wealth,df_model$EffortHealth.13,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Education,df_model$EffortHealth.13,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Income2,df_model$EffortHealth.13,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Age,df_model$EffortHealth.13,use="pairwise.complete.obs",method='spearman')
df_model <- df_model [df_model$Sex %in% c("Male", "Female"), ]
wilcox.test(df_model$EffortHealth.13~df_model$Sex)

set.seed(111297)
df_model <- merge(df,df_health,by=c('User'))
df_model$HealthBehaviorNormal <- bestNormalize(df_model$EffortHealth.13)$x.t
df_model$PUMRJulyNormal <- bestNormalize(df_model$PUMRJuly)$x.t
df_model$AgeNormal <- bestNormalize(df_model$Age)$x.t
df_model$WealthNormal <- bestNormalize(df_model$Wealth)$x.t

# Best Normalize has very similar results for Yeo Johnson and Order Norm transformation, so the chosen method is dependent
# on the seed used. We report analyze both normalization techniques in the text.
df_model$PUMRJulyNormal1 <- yeojohnson(df_model$PUMRJuly)$x.t
df_model$PUMRJulyNormal2 <- orderNorm(df_model$PUMRJuly)$x.t
risk_mixed_normal = lm(HealthBehaviorNormal~ PUMRJulyNormal + WealthNormal +  AgeNormal + Education, data = df_model)
print(summary(risk_mixed_normal))
risk_mixed_normal = lm(HealthBehaviorNormal~ PUMRJulyNormal1 + WealthNormal +  AgeNormal + Education, data = df_model)
print(summary(risk_mixed_normal))
risk_mixed_normal = lm(HealthBehaviorNormal~ PUMRJulyNormal2 + WealthNormal +  AgeNormal + Education, data = df_model)
print(summary(risk_mixed_normal))
(BNobject <- bestNormalize(df_model$PUMRJuly))

```





# Risk from air pollution, contaminated water, and occupational risk SEV and Perceived mortality risk from each cause
OLS regressions looking at these relationships

``` {r Supplemental Method 2}
# Air Pollution -------------------------------------------------
df_model <- merge(df,df_sev[df_sev[,'Risk'] == 'Air pollution',],by=c('User'))

print(paste('SEV Vals -- ',good_summary(df_model$X5_year_range)))
print(paste('PUMR AIR Pollution -- ',good_summary(df_model$PUMRAir))) 

# PUMR has high kurtosis -- makes sense because most people have low perceived probability that they will die of this cause in US. Also skewed for this reason. So we do not have a normal distribution. Definitely will want to check assumptions for model.

# Test which of the control variables are significantly related to the outcome variable
cor.test(df_model$Wealth,df_model$PUMRAir,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Education,df_model$PUMRAir,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Income,df_model$PUMRAir,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Age,df_model$PUMRAir,use="pairwise.complete.obs",method='spearman')
df_model <- df_model [df_model$Sex %in% c("Male", "Female"), ]
wilcox.test(df_model$PUMRAir~df_model$Sex)

# Looks like Age and Sex are the significant variables here
# Normalize all linear (and ordinal) independent predictors
df_model$PUMRAirNormal <- bestNormalize(df_model$PUMRAir)$x.t
df_model$X5_year_rangeNormal <- bestNormalize(df_model$X5_year_range)$x.t
df_model$AgeNormal <- bestNormalize(df_model$Age)$x.t

# 5 year and Sum to 79 -- maybe do both to look for differences, Then we can talk about if people are summing up more info (whole life) or only thinking about here and now? Supplemental analysis! 
risk_mixed_normal = lmer(PUMRAirNormal ~ X5_year_rangeNormal + AgeNormal + Sex + (1 | State_Name), data = df_model)
summary(risk_mixed_normal, diagnostics=TRUE)

risk_normal = lm(PUMRAirNormal ~ X5_year_rangeNormal + AgeNormal + Sex, data = df_model) 
summary(risk_normal,diagnostics=TRUE)




# Unsafe water  -------------------------------------------------
df_model <- merge(df,df_sev[df_sev[,'Risk'] == 'Unsafe water source',],by=c('User'))
print(paste('SEV Vals -- ',good_summary(df_model$X5_year_range)))
print(paste('PUMR Water -- ',good_summary(df_model$PUMRWater)))

# Test which of the control variables are significantly related to the outcome variable
cor.test(df_model$Wealth,df_model$PUMRWater,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Education,df_model$PUMRWater,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Income,df_model$PUMRWater,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Age,df_model$PUMRWater,use="pairwise.complete.obs",method='spearman')
df_model <- df_model [df_model$Sex %in% c("Male", "Female"), ]
wilcox.test(df_model$PUMRWater~df_model$Sex)

# Normalize age & sex (the significant controls), PUMR, and SEV
df_model$PUMRWaterNormal <- bestNormalize(df_model$PUMRWater)$x.t
df_model$X5_year_rangeNormal <- bestNormalize(df_model$X5_year_range)$x.t
df_model$AgeNormal <- bestNormalize(df_model$Age)$x.t

risk_normal = lm(PUMRWaterNormal ~ X5_year_rangeNormal + AgeNormal + Sex, data = df_model) 
summary(risk_normal,diagnostics=TRUE)




# Occupational Risk -------------------------------------------------
df_model <- merge(df,df_sev[df_sev[,'Risk'] == 'Occupational risks',],by=c('User'))
print(paste('SEV Vals -- ',good_summary(df_model$X5_year_range)))
print(paste('PUMR Occupational Risks -- ',good_summary(df_model$PUMROccupation)))

# Test which of the control variables are significantly related to the outcome variable
cor.test(df_model$Wealth,df_model$PUMROccupation,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Education,df_model$PUMROccupation,,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Income,df_model$PUMROccupation,,use="pairwise.complete.obs",method='spearman')
cor.test(df_model$Age,df_model$PUMROccupation,,use="pairwise.complete.obs",method='spearman')
df_model <- df_model [df_model$Sex %in% c("Male", "Female"), ]
wilcox.test(df_model$PUMROccupation~df_model$Sex)

# Normalize age (the significant control), PUMR, and SEV
df_model$PUMROccupationNormal <- bestNormalize(df_model$PUMROccupation)$x.t
df_model$X5_year_rangeNormal <- bestNormalize(df_model$X5_year_range)$x.t
df_model$AgeNormal <- bestNormalize(df_model$Age)$x.t

risk_normal = lm(PUMROccupationNormal ~ X5_year_rangeNormal + AgeNormal, data = df_model) 
summary(risk_normal,diagnostics=TRUE)

```



```{r}
PUMR <- c()
Month <- c()
for (i in df$PUMRMarch){
  PUMR <- c(PUMR,i)
  Month <- c(Month,'March')
}

for (i in df$PUMRMay){
  PUMR <- c(PUMR,i)
  Month <- c(Month,'May')
}

for (i in df$PUMRJuly){
  PUMR <- c(PUMR,i)
  Month <- c(Month,'July')
}

df_hist <- data.frame(PUMR=PUMR,Month=Month)

ggplot(df_hist,aes(x=PUMR,group=Month,colour=Month,fill=Month)) +
  geom_histogram(aes(y=..density..), colour="black",fill="white")+
  geom_density(alpha=0.2) +
  facet_grid(cols=vars(Month)) +
  theme_bw()

```
